---
title: "Bayesian Structure learning with ordinal data"
author: "Linduni Rodrigo"
date: "2024-09-10"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: false
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F )
```

## Setting the working directory

```{r}
#setwd("LSAC/OSEM")
```

## Loading Required Packages

```{r, message=FALSE}
library(BiDAG)
library(Rgraphviz)
library(pcalg)
library(graph)
library(bnlearn)
library(MXM)
library(sbgcop)
library(infotheo)
library(rpart)
library(devtools)
library(readr) #read csv
library(ggplot2)
library(pROC)  # for ROC curve and AUC calculation
source_url("https://raw.githubusercontent.com/cuiruifei/CausalMissingValues/master/R/inferCopulaModel.R")
```

## Source external R scripts

```{r}
# Major file containing the OSEM algorithm
source("ordinalScore.R") 

# Modified version of the existing functions in the BiDAG package
insertSource("spacefns.R",package = "BiDAG")
insertSource("usrscorefns.R",package = "BiDAG")
insertSource("initpar.R",package = "BiDAG")
insertSource("scoreagainstdag.R",package = "BiDAG")
```

# Structure learning with simulated data

## Data simulation

```{r True_DAGs}
set.seed(222)

N <- 500
n <- 10
# Generate a regular DAGs with n nodes with 4 number of neighbors
trueDAG <- randDAG(n = n, d = 4, method = "er", wFUN = list(mywFUN))


```

```{r Plot_true_DAG}
# Plot true DAG structures
plot(trueDAG, main=paste0("True DAG with ", n, " nodes"))


```


```{r Generate_Gaussian}
## generate 500 observations from the DAG using standard normal error distribution
set.seed(222)
gaussian_data_n10 <- rmvDAG2(N, trueDAG_n10)

```


```{r Gaussian_to_Ordinal}
set.seed(222)
# Convert the Gaussian dataset into an ordinal dataset
ordinal_data <- generateOrdinal(N, n[1], trueDAG_n10, exp_levels = 4, concent_param = 2)
ordinal_data_df <- as.data.frame(ordinal_data_n10)
ordinal_data_df[] <- lapply(ordinal_data_n10_df[], as.ordered)
```


```{r #levels}
# No. of levels in each variable 
(ordinal_levels <- apply(ordinal_data, 2, function(x) length(unique(x))))
```


## DAG structure learning using OSEM algorithm

```{r run_OSEM, results='hide', cache = TRUE}
# the OSEM algorithm
set.seed(123)
OSEMfit <- ordinalStructEM(n, ordinal_data,
                           usrpar = list(penType = "other",
                                         L = 5,
                                         lambda = 8,
                                         preLevels = NULL))
```

```{r Estimated_DAG_OSEM, echo=FALSE}
# Plot OSEM estimated vs true DAG structure
par(mfrow = c(1,2))
plot(trueDAG, main = "True DAG with 10 nodes")
tempgraph <- graphAM(as.matrix(OSEMfit$DAG), edgemode = "directed")
plot(tempgraph, main= "OSEM estiamted DAG")
```


## DAG structure learning using hybrid method with the BDe score and NPC as initial search space (BDe) - binary data

### Binarise data

```{r Binarise_Ordinal}
# Convert the ordianl dataset into binary dataset

# Function to binarize ordered factors based on a threshold
binarize_factor <- function(col) {
    l <- length(unique(col))
    if (l ==2){
      return(as.numeric(col))
    }else{
       # Convert ordered factors to binary values based on the threshold
       binary_var <- ifelse(as.numeric(col) > l/2, 1, 0)
       return(as.numeric(binary_var))
    }
   
}

binary_data <- as.matrix(apply(ordinal_data_df,2,binarize_factor))

# View the result
print(head(binary_data))

```

### Learn DAG structure with BDe 

```{r run_BDe, results='hide', cache = TRUE}
set.seed(123)
# hybrid method with the BDe score and the nominal PC output as the initial search space
BDE <- scoreparameters("bdecat",data.frame(binary_data),bdecatpar = list(chi = 0.5))
BDEfit <- iterativeMCMC(BDE)
```

```{r Estimated_DAG_BDe, echo = FALSE}
# Plot BDe estimated vs true DAG structure for 10 nodes
par(mfrow = c(1,2))
plot(trueDAG, main = "True DAG with 10 nodes")
tempgraph <- graphAM(as.matrix(BDEfit$DAG), edgemode = "directed")
plot(tempgraph, main= "BDe estimated DAG")
```

### Evaluating the performance

#### Calculate performance measures (SHD, Precision,TPR and FPR)

```{r Compare_TPR&FPR, message=FALSE}
library(knitr)
library(kableExtra)

OSEMfit = OSEMfit
BDEfit = BDEfit
trueDAG = trueDAG

# Calculate "SHD", "Precision","TPR","FPR_N","FPR_P"
data_frame <- data.frame(rbind(comparePatterns(as.matrix(OSEMfit$DAG),trueDAG,hardP2P = FALSE), comparePatterns(as.matrix(BDEfit$DAG),trueDAG,hardP2P = FALSE)))
row.names(data_frame) <- c("OSEM","BDe")

# Use kable to create a Markdown table
kable(data_frame[,c(1,6:9)], col.names = c("SHD", "Precision","TPR","FPR_N","FPR_P"), 
      caption = "Table of TPR and FPR for structure learned from different methods", 
      row.names = T)%>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(1, width = "3cm") %>%  
  column_spec(2, width = "4cm") 
```

#### Receiver Operating Characteristic (ROC) curve. 

Calculate True Positive Rate (TPR) and the False Positive Rate
(FPR) for the DAG structures estimated by the OSEM and BDe
methods, each employing following penalization parameters.

-   For BDe 15 different equivalent sample sizes ($\chi$)
    $\chi ∈ {0.000001, 0.00001,0.0001, 0.001, 0.01, 0.1, 1, 10, 20, 40, 60, 80, 100, 120,150}$
    and
-   For OSEM 15 different coefficient to be multiplied by the BIC
    penalty ($\lambda$)
    $\lambda \in {0.1,0.5,1, 1.5, 2, 2.5, 3, 4, 6, 10, 20, 30,40,50,60}$

```{r OSEM-Comparison, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}

trueDAG = trueDAG
data = ordinal_data
# Define parameters
lambda_values <- c(0.1,0.5,1, 1.5, 2, 2.5, 3, 4, 6, 10, 20, 30, 40, 50, 60)

tprOSEM_values <- rep(NA, nrow = length(lambda_values))
fdrpOSEM_values <- rep(NA, nrow = length(lambda_values))


# Loop through lambda values for OSEM
for (i in seq_along(lambda_values)) {
  lambda <- lambda_values[i]
  
  # Fit the OSEM model
  OSEMfit <- ordinalStructEM(n,data,
                             usrpar = list(penType = "BIC",
                                           L = 5,
                                           lambda = lambda,
                                           preLevels = NULL))
  
  # Calculate TPR and FPR
  results_soft <- comparePatterns(as.matrix(OSEMfit$DAG), trueDAG, hardP2P = FALSE)
  tprOSEM_values[i] <- results_soft[7] # TPR
  fdrpOSEM_values[i] <- results_soft[9] # FDRp
}


```

```{r BDE-Comparison, results='hide'}

trueDAG = trueDAG_n20
data = binary_data_n20
# Define parameters

chi_values <- c(0.00000001,0.0000001, 0.000001, 0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 20, 40, 60, 80, 100, 120, 150)

tprBDe_values <- rep(NA, nrow = length(chi_values))
fdrpBDe_values <- rep(NA, nrow = length(chi_values))

# Loop through chi values for BDE
for (i in seq_along(chi_values)) {
  chi <- chi_values[i]
  
  # Fit the BDE model
  BDEfit <- iterativeMCMC(scoreparameters("bdecat", data.frame(data), bdecatpar = list(chi = chi)))
  
  # Calculate TPR and FPR
  results_soft <- comparePatterns(as.matrix(BDEfit$DAG), trueDAG,hardP2P = FALSE)
  tprBDe_values[i] <- results_soft[7]
  fdrpBDe_values[i] <- results_soft[9]
}

```

```{r df_performace_matrices}


# Combine both datasets
roc_data <- rbind(data.frame(FPR = fdrpOSEM_values,
                           TPR = tprOSEM_values,
                           Method = "OSEM"), data.frame(FPR = fdrpBDe_values,
                           TPR = tprBDe_values,
                           Method = "BDe"))


```

```{r ROC, echo = FALSE}

# Plot ROC curves for both methods
ggplot(roc_data, aes(x = FPR, y = TPR, color = Method)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(x = "False Positive Rate", y = "True Positive Rate", title = "ROC Curve Comparison") +
  theme_minimal() + ylim(0.1,1)+xlim(0.1,1)+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = c("blue", "red"))

```

